Supplemental Materials for: \\ Large scale risks from pesticides in small streams

### 2017-05-01 article.tex
2nd round of revisions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Eduard Szöcs}
\affiliation[Institute for Environmental Sciences]{Institute for Environmental Sciences, University of Koblenz-Landau, Germany}
\email{szoecs@uni-landau.de}
\phone{+49 (0)6341 280 31552}

@@

\author{Ralf B. Schäfer}
\affiliation[University Koblenz-Landau]{Institute for Environmental Sciences, University of Koblenz-Landau, Germany}

@@

\begin{abstract}
% 150-200 words
Small streams are important refugia for biodiversity.
In agricultural areas they may be at risk from pesticide pollution. 
However, most related studies have been limited to a few streams on the regional level, hampering extrapolation to larger scales. 
We quantified risks as exceedances of regulatory acceptable concentrations (RACs) and used German monitoring data to quantify the drivers thereof and to assess current risks in small streams on a large scale. 
The data set comprised of 1,766,104 measurements of 478 pesticides (including metabolites) related to 24,743 samples from 2,301 sampling sites. 
We investigated the influence of agricultural land use, catchment size, as well as precipitation and seasonal dynamics on pesticide risk taking also concentrations below the limit of quantification into account. 
The exceedances of risk thresholds dropped 3.7-fold at sites with no agriculture, indicating that agricultural land use is a major contributor of pesticides in streams.
Precipitation increased detection probability by 43\% and concentrations were the highest from April to June. 
RACs were exceeded in 26\% of streams, with the highest exceedances found for neonicotinoid insecticides. 
We conclude that pesticides from agricultural land use are a major threat to small streams and their biodiversity. 
To reflect peak concentrations, current pesticide monitoring needs refinement.

@@

The aim of this study is to identify drivers and dynamics of pesticide concentrations in streams on large spatial scales.
To achieve this, we compiled and analysed large-scale pesticide monitoring data from small streams in Germany and examined four hypotheses:
1) A major fraction of pesticides is applied to agricultural fields.
Therefore, we hypothesised that the most frequent exceedances of RACs occur in streams with a high proportion of agricultural land use in the catchment. 
If agricultural land use was indeed the main source for pesticides in streams, the RAC exceedances should drop to negligible levels in the absence of agricultural land use in their catchments. 
Given this drop we expected a non-linear relationship between exceedances and agricultural land use.
In case of a non-linear relationship, our analyses might guide the definition of reference streams without pesticide pollution in future monitoring. 
2) We hypothesised that an increase in stream size is associated with an decrease in RAC exceedances, given a higher dilution potential. 
3) However, also the timing of sampling may influence measured concentrations,
as local and regional studies reported higher pesticide concentrations after precipitation events \citep{liess_determination_1999,Xing_Chow_Rees_Meng_Li_Ernst_Benoy_Zha_Hewitt_2013}. 
Therefore, we hypothesised the highest RAC exceedances to be found after precipitation events. 
4) Pesticides are not applied throughout the whole year and highest RAC exceedances should be found during the main growing season.
Finally, we quantified the current risks from pesticides in small streams in Germany and the compounds accountable for the risk.

@@

Given their high abundance in the landscape \citep{nadeau_hydrological_2007} small streams below 10~km\textsuperscript{2} are disproportionally less sampled in current monitoring (Figure~\ref{fig:fig3}), which may be attributed to the missing categorisation in the WFD. 
Clearly, there is currently a lack of knowledge on stressor effects on small streams.
We analysed only data from small streams, however, for lentic small water bodies this lack might be even greater \citep{lorenz_specifics_2016}.

@@

% size
We did not find a statistically significant relationship between pesticide pollution and catchment size (Hypothesis 2). 
However, previous studies showed that small streams are more polluted than bigger streams \citep{schulz_field_2004,stehle_pesticide_2015,knauer_pesticides_2016}.
This can be explained by the relatively short gradient of catchment sizes in our dataset, with most of the streams with catchments above $10~km^2$ and below $100~km^2$ (Figure~\ref{fig:fig3}, top).
For example, the gradient of \citet{schulz_field_2004} covered 6 orders of magnitude.

@@

We found the highest the probability of exceeding LOQ from April to June (10\% for Q2) and lowest in the first quarter of the year (4\%, Figure~\ref{fig:fig5}, bottom right).
This annual pattern coincides, as hypothesised, with the main application season for pesticides in Central Europe.
Nevertheless, there are compound-specific differences in the annual pattern, which explains the wide CI for the absolute RQ (Figure~\ref{fig:fig5}, bottom left). 
For example, the herbicide Diflufenican showed the highest RQ and the highest probability of exceeding LOQ during the winter quarters Q1 and Q4 (Supplemental Table~S4), which coincides with the application period it is registered for in Germany \citep{bvl_online_2016}.
Moreover, compound properties, like half-life or water solubility, might influence compound dynamics. 
Our study suggests that pesticide risks display compound specific spatio-temporal dynamics. 
Currently, little is known about these and further research on those might provide useful information for future environmental risk assessment. 
For example, the sensitivity of organisms is often life stage dependent \citep{hutchinson1998analysis} and knowledge on temporal dynamics could inform on concurrent exposure to multiple pesticides, as well as assist to parameterise toxicokinetic and toxicodynamic models \citep{ashauer2016modelling}. 
Moreover, our results show that analysing absolute concentrations and probabilities of LOQ together might deliver valuable insights into risk dynamics.

@@

% Approval / Risk Assessment
Monitoring data, despite the outlined limitations, provides an opportunity to study large-scale environmental occurrence patterns of pesticides.
Furthermore, such nationwide compilations, may not only be used for governmental surveillance, but also to answer other questions, like validation of exposure modelling \cite{knabel_fungicide_2014}, retrospective evaluation of regulatory risk assessment \citep{knauer_pesticides_2016,stehle_pesticide_2015}or occurrences of pesticide mixtures \cite{schreiner_pesticide_2016}.
However, the sampling design needs to account for precipitation events to provide robust data. 
Our results suggest that exceedances of RACs are landscape dependent % = Zusammenhang RAC<->Agriculture
and therefore, pesticide regulation should account for landscape features. 
Moreover, the high exceedances of RACs indicate that greater efforts are needed to describe causal links, which may lead to further developments of the current authorisation procedure.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Eduard Szöcs}
\affiliation[University Koblenz-Landau]{Institute for Environmental Sciences, University of Koblenz-Landau, Fortstraße 7, 76829 Landau, Germany}

@@

\author{Ralf B. Schäfer}
\affiliation[University Koblenz-Landau]{Institute for Environmental Sciences, University of Koblenz-Landau, Fortstraße 7, 76829 Landau, Germany}
\email{schaefer-ralf@uni-landau.de}
\phone{+49 (0)6341 280 31536}
\fax{+49 (0)6341 280 31326}

@@

\begin{abstract}
% 150-200 words
Small streams are important refugia for biodiversity.
In agricultural areas they may be at risk from pesticide pollution. 
However, most related studies have been limited to a few streams on the regional level, hampering extrapolation to larger scales. 
We quantified risks as exceedances of regulatory acceptable concentrations (RACs) and used German monitoring data to quantify the drivers thereof and to assess current risks in small streams on a large scale. 
The data set comprised of 1,766,104 measurements of 478 pesticides (including metabolites) related to 24,743 samples from 2,301 sampling sites. 
We investigated the influence of agricultural land use, catchment size, as well as precipitation and seasonal dynamics on pesticide risk taking also concentrations below the limit of quantification into account. 
The exceedances of risk thresholds dropped 3.7-fold at sites with no agriculture.
Precipitation increased detection probability by 43\% and concentrations were the highest from April to June. 
Overall, this indicates that agricultural land use is a major contributor of pesticides in streams.
RACs were exceeded in 26\% of streams, with the highest exceedances found for neonicotinoid insecticides. 
We conclude that pesticides from agricultural land use are a major threat to small streams and their biodiversity. 
To reflect peak concentrations, current pesticide monitoring needs refinement.

@@

The aim of this study is to identify drivers and dynamics of pesticide concentrations in streams on large spatial scales.
To achieve this, we compiled and analysed large-scale pesticide monitoring data from small streams in Germany and examined four hypotheses:
1) A major fraction of pesticides is applied to agricultural fields.
Therefore, we hypothesised that the most frequent exceedances of RACs occur in streams with a high proportion of agricultural land use in the catchment. 
If agricultural land use was indeed the main source for pesticides in streams, the RAC exceedances should drop to negligible levels in the absence of agricultural land use in their catchments. 
Given this possible drop we expected a non-linear relationship between exceedances and agricultural land use.
In case of a non-linear relationship, our analyses might guide the definition of reference streams without pesticide pollution in future monitoring. 
2) Based on previouse studies, we hypothesised that an increase in catchment size is associated with an decrease in RAC exceedances \citep{stehle_pesticide_2015,schulz_field_2004}. 
3) However, also the timing of sampling may influence measured concentrations,
as local and regional studies reported higher pesticide concentrations after precipitation events \citep{liess_determination_1999,Xing_Chow_Rees_Meng_Li_Ernst_Benoy_Zha_Hewitt_2013}. 
Therefore, we hypothesised the highest RAC exceedances to be found after precipitation events. 
4) Pesticides are not applied throughout the whole year and highest RAC exceedances should be found during the main growing season.
Finally, we quantified the current risks from pesticides in small streams in Germany and the compounds accountable for the risk.

@@

Given their high abundance in the landscape \citep{nadeau_hydrological_2007} small streams below 10~km\textsuperscript{2} are disproportionally less sampled in current monitoring (Figure~\ref{fig:fig3}).
This may be attributed to the missing categorisation in the WFD, but also to technical (e.g. sampling at low flow) and natural reasons (e.g. emphemeral streams).

Clearly, there is currently a lack of knowledge on stressor effects on small streams.
We analysed only data from small streams, however, for lentic small water bodies this lack might be even greater \citep{lorenz_specifics_2016}.

@@

% size
We did not find a statistically significant relationship between pesticide pollution and catchment size (Hypothesis 2). 
However, previous studies showed that small streams are more polluted than bigger streams \citep{schulz_field_2004,stehle_pesticide_2015,knauer_pesticides_2016}.
This can be explained by the relatively short gradient of catchment sizes in our dataset, with most of the streams with catchments above $10~km^2$ and below $100~km^2$ (Figure~\ref{fig:fig3}, top).
For example, the gradient of \citet{schulz_field_2004} covered 6 orders of magnitude. 
Although there is a relationship between catchment size and stream size (Supplemental Figure~S11), this might be also influenced by geology and precipitation regime. 
Therefore, our results cannot be directly translated to stream size.

@@

We found the highest the probability of exceeding LOQ from April to June (10\% for Q2) and lowest in the first quarter of the year (4\%, Figure~\ref{fig:fig5}, bottom right).
This annual pattern coincides, as hypothesised, with the main application season for pesticides in Central Europe.
Nevertheless, there are compound-specific differences in the annual pattern, which explains the wide CI for the absolute RQ (Figure~\ref{fig:fig5}, bottom left). 
For example, the herbicide Diflufenican showed the highest RQ and the highest probability of exceeding LOQ during the winter quarters Q1 and Q4 (Supplemental Table~S4), which coincides with the application period it is registered for in Germany \citep{bvl_online_2016}.
Moreover, compound properties, like half-life or water solubility, might influence compound dynamics. 
Our study suggests that pesticide risks display compound specific spatio-temporal dynamics. 
Currently, little is known about these and further research on those might provide useful information for future environmental risk assessment. 
For example, the sensitivity of organisms is often life stage dependent \citep{hutchinson1998analysis} and knowledge on temporal dynamics could inform on concurrent exposure to multiple pesticides, as well as assist to parameterise toxicokinetic and toxicodynamic models \citep{ashauer2016modelling}. 
Moreover, our results show that analysing absolute concentrations and probabilities of LOQ together might deliver valuable insights into risk dynamics.
The influence of agricultural land use within the catchment area and the coincidence with the growing season indicates that agricultural land use a major contributor of pesticides in streams.

@@

% Approval / Risk Assessment
Monitoring data, despite the outlined limitations, provides an opportunity to study large-scale environmental occurrence patterns of pesticides.
Furthermore, such nationwide compilations, may not only be used for governmental surveillance, but also to answer other questions, like validation of exposure modelling \cite{knabel_fungicide_2014}, retrospective evaluation of regulatory risk assessment \citep{knauer_pesticides_2016,stehle_pesticide_2015} or occurrences of pesticide mixtures \cite{schreiner_pesticide_2016}.
However, the sampling design needs to account for precipitation events to provide robust data. 
Therefore, non-linear modeling can provide additional insights to risk assessment compared to linear modeling \citep{fewster_analysis_2000}. 
Our results suggest that exceedances of RACs are landscape dependent and therefore, pesticide regulation should account for landscape features. 
Moreover, the high exceedances of RACs indicate that greater efforts are needed to describe causal links, which may lead to further developments of the current authorisation procedure.

### 2017-05-01 supplement/Supplemental_Materials.tex
2nd round of revisions

\vfill
\begin{description}
	\item[Number of pages: ] \totalpages
	\item[Number of figures: ] \totalfigures
	\item[Number of tables: ] \totaltables
\end{description}
% increase page counter by 1
\setcounter{page}{2}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{fig_map}
	\caption{Spatial distribution of the 2,301 small stream sampling sites. Colour codes different federal states (see Supplemental Table S1 for abbreviations). Overlayed are ecoregions. ER4: Alps (no label), ER8: Western highlands,
	ER9: Central highlands, ER 13: Western plains , ER14: Central plains. }
	\label{fig:ecoreg}
\end{figure}

@@

%% -------------------------------------------------------------------------
\chapter{Influence of agricultural land use and catchment size}
\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{ezgagrirac}
	\caption[Raw data used for the model in equation 2 and Figure 3 of the main article.]{Raw data used for the model in equation 2 and Figure 3 of the main article. Color codes the number of RAC exceedances (on a log-scale). Grey points denote sites without any exceedance.}
	\label{fig:ezgagrirac}
\end{figure}

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.8\textwidth]{mean_coef.pdf}
  \caption{Summarised coefficients (and their 95\% CI) for precipitation (top row) and quarter of the year (bottom row) from a meta-analysis of the 22 modelled compounds (see Figure S6). Left: coefficients for mean RQ ($\mu$), right: coefficients for probability of exceeding LOQ ($\nu$). 
  Coefficients are shown on the link scale.
  Single compound coefficients are provided in Supplemental Table~S4 and Figure~S7.
  }
  \label{fig:mean_coefs}
\end{figure}

@@

% \bibliographystyle{apalike}      % basic style, author-year citations
% \bibliography{references}

>>>>

\vfill
\begin{description}
	\item[Number of pages: ] \totalpages
	\item[Number of figures: ] \totalfigures
	\item[Number of tables: ] \totaltables
\end{description}
% increase page counter by 1
\setcounter{page}{1}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{fig_map}
	\caption[Spatial distribution of the 2,301 small stream sampling sites.]{Spatial distribution of the 2,301 small stream sampling sites. Colour codes different federal states (see Supplemental Table S1 for abbreviations). Overlayed are ecoregions. ER4: Alps (no label), ER8: Western highlands,
	ER9: Central highlands, ER 13: Western plains , ER14: Central plains. }
	\label{fig:ecoreg}
\end{figure}

@@

%% -------------------------------------------------------------------------
\chapter{Influence of agricultural land use and catchment size}

\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{ezgagrirac}
	\caption[Raw data used for the model in Eqn. 2 and Fig. 3 of the main article.]{Raw data used for the model in equation 2 and Figure 3 of the main article. Color codes the number of RAC exceedances (on a log-scale). Grey points denote sites without any exceedance.}
	\label{fig:ezgagrirac}
\end{figure}

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.8\textwidth]{mean_coef.pdf}
  \caption[Summarised coefficients (and their 95\% CI) for precipitation and quarter of the year]{Summarised coefficients (and their 95\% CI) for precipitation (top row) and quarter of the year (bottom row) from a meta-analysis of the 22 modelled compounds (see Figure S6). Left: coefficients for mean RQ ($\mu$), right: coefficients for probability of exceeding LOQ ($\nu$). 
  Coefficients are shown on the link scale.
  Single compound coefficients are provided in Supplemental Table~S4 and Figure~S7.
  }
  \label{fig:mean_coefs}
\end{figure}

@@

%% -------------------------------------------------------------------------
\chapter{Additional material}

\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{width_size}
	\caption[Relationship between catchment size and stream width.]{Relationship between catchment size and stream width. A power function has been fitted to each dataset separately (colored dashed lines) and the combined dataset (black line and equation). The gray rectangle marks the estimated catchment size for a width of 1 m (=7 km\textsuperscript{2}). Data has been compiled from three different sources (\citet{Fern2015}, \citet{Voss2015} and the federal state of Thuringia)}
	\label{fig:size_width}
\end{figure}

\bibliographystyle{apalike}      % basic style, author-year citations
\bibliography{references}
